---
title: "Analysis of pneumococcal deaths in the US"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)
library(dplyr)
library(lubridate)
library(ggplot2)
library(MASS)
library(arules)
library(usdata)
library(covidcast)
library(arulesViz)
source('./R/death_plot_fun.R')
source('./R/pred_interval.R')

source('./R/Format_US_Population.R')

#https://github.com/socdataR/wonderapi
#devtools::install_github("socdataR/wonderapi", build_vignettes = F, force=T)
#library(wonderapi)

```

```{r}
pneumo.deaths <-   readRDS('./CONFIDENTIAL/pneumococcal_deaths_line_list_geo.rds')

#the file has other suspected deaths like lobar pneumonia; this filters those out
pneumo.deaths <- pneumo.deaths %>%
 # filter(pneumococcal==1 | j181==1 | j154==1)
  filter(pneumococcal==1 )

pneumo.deaths$agec <- factor(pneumo.deaths$agec, c("Under 5 Years",'5-24 years', "25-44 years","45-64 years","65-74 years" ,"75-84 years","85 years and older"))

pneumo.deaths$race_recode <- factor(pneumo.deaths$race_recode, levels=c(1,2,3,4,5), c('White','Black','Hispanic','American Indian', 'Asian/Pacific Islanders')) 

pneumo.deaths$one <- 1

#Collect all of the ICD codes into a single field
pneumo.deaths$all_icd <- apply(pneumo.deaths[, grep('icd', names(pneumo.deaths))], 1, function(x) paste(x, collapse=' ' ))

pneumo.deaths$J13 <- grepl('J13',pneumo.deaths$all_icd)
pneumo.deaths$A403 <- grepl('A403',pneumo.deaths$all_icd)

pneumo.deaths$pneumo <- 1

table(pneumo.deaths$J13,pneumo.deaths$A403 )
```


Popsize data
```{r}
#Format_US_Population()
pop1 <- readRDS('./Data/pop_interpol.rds')

pop2 <- pop1 %>%
  group_by(date, agec, race_recode) %>%
  summarize(pop.interpol= sum(pop.interpol)) %>%
  mutate(race_recode= factor(race_recode,levels=c(1,2,3,4,5), c('White','Black','Hispanic','American Indian', 'Asian/Pacific Islanders')))
```

Deaths by state
```{r}
state_deaths <- pneumo.deaths %>%
  filter(year==2019) %>%
  group_by(state) %>%
  summarize(N_deaths=n())

state_pop <- state_stats %>%
  dplyr::select(abbr, pop2010) %>%
  mutate(state=as.character(abbr))

state_deaths <- merge (state_deaths, state_pop, by='state') %>%
  mutate(inc= N_deaths/pop2010*100000 )
```

Deaths by ABCs regions https://www.cdc.gov/abcs/methodology/surv-pop.html 
Note in 2019, there were 0.98 deaths/100,000 due to IPD in the ABCs catchment area. ABCs captures in-hospital deaths only

```{r}
state_deaths_abcs <- pneumo.deaths %>%
  filter(year==2019 & (state %in% c('CT','MN','NM'))) %>%
  group_by(state) %>%
  summarize(N_deaths=n())

state_pop <- state_stats %>%
  dplyr::select(abbr, pop2010) %>%
  mutate(state=as.character(abbr)) %>%
    rename(pop=pop2010)


state_deaths_abcs <- merge (state_deaths_abcs, state_pop, by='state') %>%
  mutate(inc= N_deaths/pop*100000 )


#name_to_fips("Miami", ties_method = "all")

#CA 3,706,404 (3 county Bay area)
CA_counties <- name_to_fips(c("Alameda", "Contra Costa","San Francisco"), state='CA', ties_method = "first")

#CO 2,835,257 (5 county Denver area)
CO_counties <- name_to_fips(c('Adams', 'Arapahoe', 'Denver', 'Douglas', 'Jefferson'), state='CO', ties_method = "first")

#GA 5,834,904 (20 county Atlanta area)  20 county metropolitan statistical area of Atlanta (found counties on slides from Monica Farley )
GA_counties <- name_to_fips(c('Cobb','Douglas','Fulton', 'Clayton', 'Dekalb','Rockdale', 'Newton','Gwinnett','Barrow', 'Walton', 'Forsyth', 'Cherokee','Pickens',"Bartow", 'Paulding','Carroll', 'Coweta','Fayette', 'Spalding', 'Henry'), state='GA', ties_method = "first")

#MD  2,749,672 (6 county Baltimore area)
MD_counties <- name_to_fips(c('Baltimore city', 'Baltimore County', 'Anne Arundel', 'Carroll', 'Harford','Howard'), state='MD', ties_method = "first")

#Ny 2,163,175 (7 county Rochester area and 8 county Albany area) 
NY_counties <- name_to_fips(c('Albany', 'Columbia', 'Genesee', 'Greene', 'Livingston', 'Monroe', 'Montgomery', 'Ontario', 'Orleans', 'Rensselaer', 'Saratoga', 'Schenectady', 'Schoharie', 'Wayne', 'Yates'), state='NY', ties_method = "first")

#Oregon 1,832,634 (3 county Portland area)
OR_counties <- name_to_fips(c("Multnomah", "Washington","Clackamas"), state='OR', ties_method = "first")

#TN 4,154,029 (20 urban counties)--note 1 county is missing here. this list is from the TN EIP site
TN_counties <- name_to_fips(c('Anderson', 'Blount', 'Cheatham', 'Davidson', 'Dickson', 'Grainger', 'Hamilton, Jefferson', 'Knox', 'Loudon', 'Madison', 'Roane', 'Robertson', 'Rutherford', 'Sevier', 'Shelby', 'Sumner', 'Union', 'Williamson', 'Wilson'), state='TN', ties_method = "first")


all.abcs.counties <- c(CA_counties,CO_counties, GA_counties,MD_counties,NY_counties,OR_counties,TN_counties)

pneumo.deaths <- pneumo.deaths %>%
  mutate( state_fips= substr(abbr_to_fips(state),1,2),
          county_fips=paste0(state_fips, county))


county_deaths_abcs <- pneumo.deaths %>%
  filter(year==2019 & (county_fips %in% all.abcs.counties)) %>%
  group_by(county_fips) %>%
  summarize(N_deaths=n())

county_pop <- county_census %>%
  dplyr::select(FIPS, POPESTIMATE2019 ) %>%
  filter(FIPS %in% all.counties) %>%
  rename(pop=POPESTIMATE2019)

county_deaths_abcs <- merge(county_deaths_abcs, county_pop, by.x='county_fips', by.y='FIPS', all=T)

county_deaths_abcs$N_deaths[is.na(county_deaths_abcs$N_deaths)] <- 0

all_abcs <- bind_rows(county_deaths_abcs,state_deaths_abcs) 

all_abcs %>%
  summarize( N_deaths=sum(N_deaths), pop=sum(pop)) %>%
  mutate(inc=N_deaths/pop*100000)


abcs.state.counties <- county_census %>%
  filter(STNAME %in% c('Connecticut', 'Minnesota', 'New Mexico')) 

#FIPS for all counties included in catchment, including full states
all.abcs.state.counties <- c(all.abcs.counties,abcs.state.counties$FIPS)
```

```{r}
sort(table(pneumo.deaths$all_icd), decreasing=T)[1:100]
```



```{r}
p1 <- death_plot_fun(year,month, J13 )
p1$plot1 +
  facet_wrap(~J13)

p1 <- death_plot_fun(year,month, A403 )
p1$plot1 +
  facet_wrap(~A403)

p1 <- death_plot_fun(year,month, j154 )
p1$plot1 +
  facet_wrap(~j154)

p1 <- death_plot_fun(year,month, j181 )
p1$plot1 +
  facet_wrap(~j181)

p1.pneumococcal <- death_plot_fun(year,month )

p1.pneumococcal$plot1
```

```{r}
p2 <- death_plot_fun(year,month, agec) 
p2$plot1 + facet_wrap(~agec, scales='free')
```

By syndrome as well
```{r, fig.height=10, fig.width=5}
p2a <- death_plot_fun(year,month, agec, J13) 
p2a$plot1 + facet_wrap(agec~J13, scales='free', ncol=2)
```


Note; big differences in age distribution. Hispanic skews much younger
```{r}
p3 <- death_plot_fun(year,month, race_recode) 
p3$plot1+ facet_wrap(~race_recode, scales='free')
```


Age of deaths, by race/ethnicity. The proportion of deaths among kids is much higher for Black and Hispanic populations. need to look at incidence by age
```{r}

p3 <- death_plot_fun(year,month, race_recode, agec) 

 p4 <- ggplot(pneumo.deaths, aes(x=agey))+
  geom_histogram() +
  theme_classic() +
  facet_wrap(~race_recode, scales='free_y')
p4
```

```{r}
p3 <- death_plot_fun(year,month, race_recode, agec) 

p3$ts$date <- as.Date(paste(p3$ts$year, p3$ts$month, '01',sep='-') , '%Y-%m-%d' )

p3$ts$year <- as.numeric(as.character(p3$ts$year))

p3.ds <- merge(p3$ts, pop2, by=c('agec','race_recode', 'date')) %>%
  filter(year<2020) %>% #Pre-padnemic rates
  group_by(race_recode, agec) %>%
  summarize(N_deaths=sum(N_deaths), pop=mean(pop.interpol))%>%
  ungroup() %>%
  mutate(inc=N_deaths/pop*100000 , race_recode=factor(race_recode)) 

ggplot(p3.ds, aes(x=agec, y=log(inc), group=race_recode,col=race_recode)) +
  geom_line()+
  theme_classic()+
  ggtitle('Rate of death by age and race/ethnicity')+
   theme(axis.text.x = element_text(angle = 45, vjust = 0.9, hjust=1)) +
  ylab('log(Deaths/100,000)') +
  xlab('')
```


## Question 1: Did the rate of death due to pneumococcus change during the pandemic?

Fit model to 2014-2019 data, extrapolate to 2020 and beyond

```{r}
ds1 <- p1.pneumococcal$ts %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-')),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         N_deaths_pre= if_else(year %in% '2020', NA_real_,N_deaths),
         log_offset=log(1))

#Poisson
mod1 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6,
            data=ds1, family='poisson')

#Negative binomial
# mod1 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6,
#             data=ds1)

ds1 <- cbind.data.frame(ds1,pred_interval(mod=mod1,model.type='poisson'))

ggplot(data=ds1, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2)


ds1 <- ds1 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds1, aes(x=date, y=rr_median)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
  ylim(0.5,2.2) +
  geom_hline(yintercept=1, lty=2, col='red')

```

Did the rate change by age group or ethnic group?
```{r}
ds2 <- p2$ts %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-')),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         N_deaths_pre= if_else(year %in% '2020', NA_real_,N_deaths),
         log_offset=log(1)) %>%
  filter( !is.na(agec) & agec %in% c('25-44 years','45-64 years','65-74 years', '75-84 years','85 years and older' ) ) 

mod.mat2 <- as.data.frame(model.matrix(~index+ sin12+cos12 +sin6 +cos6 + agec, data=ds2))
names(mod.mat2) <- gsub('-','_',names(mod.mat2))
names(mod.mat2) <- gsub(' ','_',names(mod.mat2))

ds2 <- cbind.data.frame(ds2,mod.mat2[,grep('agec', names(mod.mat2))] )

#Poisson
mod2 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6  + agec5_24_years + agec25_44_years +
                                      agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older , 
            data=ds2, family='poisson')

#Negative binomial
# mod2 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6 + agec5_24_years + agec25_44_years +
#                                       agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older, 
#             data=ds2)

ds2 <- cbind.data.frame(ds2,pred_interval(mod=mod2,model.type='poisson',ds.glm=ds2))

ggplot(data=ds2, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
  facet_wrap(~agec)+
      geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') 



ds2 <- ds2 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds2, aes(x=date, y=rr_median)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
  ylim(0.1,5) +
  geom_hline(yintercept=1, lty=2, col='red') +
    geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') +

  facet_wrap(~agec)

```

race/ethnicity and age

```{r, fig.width=8, fig.height=8}
ds3 <- p3$ts %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-')),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         N_deaths_pre= if_else(year %in% '2020', NA_real_,N_deaths),
         log_offset=log(1)) %>%
  filter( !is.na(agec) & agec %in% c('25-44 years','45-64 years','65-74 years', '75-84 years','85 years and older' ) ) 

mod.mat3 <- as.data.frame(model.matrix(~index+ sin12+cos12 +sin6 +cos6 + agec +race_recode, data=ds3))
names(mod.mat3) <- gsub('-','_',names(mod.mat3))
names(mod.mat3) <- gsub(' ','_',names(mod.mat3))
names(mod.mat3) <- gsub('/','_',names(mod.mat3))
#names(mod.mat3) <- gsub('race_recode','',names(mod.mat3), fixed=T)

ds3 <- cbind.data.frame(ds3,mod.mat3[,c(grep('agec', names(mod.mat3)), grep('race_recode', names(mod.mat3)))] )

#Poisson
mod3 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6  + agec5_24_years + agec25_44_years +
                                      agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older  + race_recodeBlack +race_recodeHispanic + race_recodeAmerican_Indian + race_recodeAsian_Pacific_Islanders , 
            data=ds3, family='poisson')

#Negative binomial
# mod2 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6 + agec5_24_years + agec25_44_years +
#                                       agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older, 
#             data=ds2)

ds3 <- cbind.data.frame(ds3,pred_interval(mod=mod3,model.type='poisson',ds.glm=ds3))

ggplot(data=ds3, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
  facet_wrap(race_recode~agec, scales='free')+
      geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') 



ds3 <- ds3 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds3, aes(x=date, y=rr_median)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
  ylim(0.1,5) +
  geom_hline(yintercept=1, lty=2, col='red') +
    geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') +

  facet_wrap(race_recode~agec)

```

Non-white race/ethnicity is very sparse--try to dichotomize race (white vs non-white)
```{r, fig.width=8, fig.height=3}
ds3 <- p3$ts %>%
  mutate( race_recode=as.character(race_recode),
         race_recode=if_else(race_recode=='White', 'White', 'non_White'),
         race_recode=as.factor(race_recode)) %>%
    group_by( year ,month, agec, race_recode) %>%
  summarize(N_deaths=sum(N_deaths)) %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-')),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         N_deaths_pre= if_else(year %in% '2020', NA_real_,N_deaths),
         log_offset=log(1)) %>%
  filter( !is.na(agec) & agec %in% c('25-44 years','45-64 years','65-74 years', '75-84 years','85 years and older' ) ) 


mod.mat3 <- as.data.frame(model.matrix(~index+ sin12+cos12 +sin6 +cos6 + agec +race_recode, data=ds3))
names(mod.mat3) <- gsub('-','_',names(mod.mat3))
names(mod.mat3) <- gsub(' ','_',names(mod.mat3))
names(mod.mat3) <- gsub('/','_',names(mod.mat3))
#names(mod.mat3) <- gsub('race_recode','',names(mod.mat3), fixed=T)

ds3 <- cbind.data.frame(ds3,mod.mat3[,c(grep('agec', names(mod.mat3)), grep('race_recode', names(mod.mat3)))] )

#Poisson
mod3 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6  + agec5_24_years + agec25_44_years +
                                      agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older  + race_recodeWhite , 
            data=ds3, family='poisson')

#Negative binomial
# mod2 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6 + agec5_24_years + agec25_44_years +
#                                       agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older, 
#             data=ds2)

ds3 <- cbind.data.frame(ds3,pred_interval(mod=mod3,model.type='poisson',ds.glm=ds3))

ggplot(data=ds3, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
  facet_wrap(race_recode~agec, scales='free', nrow=2)+
      geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') 

ggplot(data=ds3[ds3$year==2020,], aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
  facet_wrap(race_recode~agec, scales='free', nrow=2)+
      geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') 


ds3 <- ds3 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds3, aes(x=date, y=rr_median)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
  ylim(0.1,5) +
  geom_hline(yintercept=1, lty=2, col='red') +
    geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') +

  facet_wrap(race_recode~agec, nrow=2)
```

By region (for use with internal data only)

```{r}

pneumo.deaths <- pneumo.deaths %>%
  mutate(region = if_else( state %in% c('ME','VT','NH','MA','CT','RI','NY','NJ','PA'), 'Northeast',
                  if_else( state %in% c('DE','DC','FL','GA','MD','NC','SC','VA','WV','AL','MS','KY','TN','LA','OK','TX', 'AR'), 'South',
                  if_else( state %in% c('AZ','CA','CO','ID','NM','MT','UT','WY','NV','AK','CA','HI','OR','WA'), 'West', 
    if_else(state %in% c('IN','IL','MI','OH','WI','IA','KS','MN','MO','NE','ND','SD'), 'Midwest', 'NA' )))))


p3_reg <- death_plot_fun(year,month, region) 

p3_reg$plot1+ facet_wrap(~region, scales='free')
```

By ABCS vs non-ABCs catchment (for use with internal data only)

```{r}
pneumo.deaths <- pneumo.deaths %>%
  mutate(abcs_region = (county_fips %in% all.abcs.state.counties  ) )

p3_abcs <- death_plot_fun(year,month, abcs_region) 

p3_abcs$plot1+ facet_wrap(~abcs_region, scales='free')


pneumo.deaths %>%
  group_by(abcs_region, year) %>%
  summarize(N_deaths=n())
```


Occupation--86% missing

```{r}
unique(pneumo.deaths$occupation)

sum(is.na(pneumo.deaths$occupation))/nrow(pneumo.deaths)
```

Education--2.5% missing

Education (1989 revision)
00 No formal education
00-08Years of elementary school
09 1 year of high school
10 2 years of high school
11 3 years of high school
12 4 years of high school
13 1 year of college
14 2 years of college
15 3 years of college
16 4 years of college
17 5 or more years of college
99 Not stated 

Education (2003 revision)
Field is blank for registration areas that are using the 1989 revision format of the item.
1 ... 8th grade or less
2 ... 9 - 12th grade, no diploma
3 ... high school graduate or GED completed
4 ... some college credit, but no degree
5 ... Associate degree
6 ... Bachelor’s degree
7 ... Master’s degree
8 ... Doctorate or professional degree
9 ... Unknown 

Education reporting flag
0 1989 revision of education item on certificate
1 2003 revision of education item on certificate
2 no education item on certificate

education_recode: 1=HS or less; 2= some college or more; 99=missing

```{r, fig.width=6, fig.height=10}
unique(pneumo.deaths$education1989)

unique(pneumo.deaths$education2003)

unique(pneumo.deaths$education_flag)

pneumo.deaths <- pneumo.deaths %>%
  mutate(education_hs = if_else(education1989 %in% c('00','01','02','03','04','05','06','07','08','09','10','11','12') |education2003 %in% c('1','2','3') ,1,0 ),
         education_college = ifelse( education1989 %in% c('13','14','15','16','17') | education2003 %in% c('4','5','6','7','8'),1,0),
         education_missing= ifelse( education1989 %in% c('99') | education2003 %in% c('9'),1,0),
         education_recode = ifelse(education_hs==1,1, ifelse(education_college==1,2,99) )
         )

table(pneumo.deaths$education_recode)

p5 <- death_plot_fun(year,month,agec, education_recode) 
p5$plot1+ facet_wrap(agec~education_recode, scales='free', ncol=3)
```

This shows a clear spike in pneumococcal mortality among young adults with HS education or lower during the first wave. This makes sense since these people are probably the least likely to be able to isolate
```{r}

ds5 <- p5$ts %>%
  filter(education_recode %in% c(1,2)) %>%
  mutate( education_recode=as.factor(education_recode)) %>%
    group_by( year ,month, agec, education_recode) %>%
  summarize(N_deaths=sum(N_deaths)) %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-')),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         N_deaths_pre= if_else(year %in% '2020', NA_real_,N_deaths),
         log_offset=log(1)) %>%
  filter( !is.na(agec) & agec %in% c('25-44 years','45-64 years','65-74 years', '75-84 years','85 years and older' ) ) 


mod.mat5 <- as.data.frame(model.matrix(~index+ sin12+cos12 +sin6 +cos6 + agec +education_recode, data=ds5))
names(mod.mat5) <- gsub('-','_',names(mod.mat5))
names(mod.mat5) <- gsub(' ','_',names(mod.mat5))
names(mod.mat5) <- gsub('/','_',names(mod.mat5))
#names(mod.mat5) <- gsub('education_recode','',names(mod.mat5), fixed=T)

ds5 <- cbind.data.frame(ds5,mod.mat5[,c(grep('agec', names(mod.mat5)), grep('education_recode', names(mod.mat5)))] )

#Poisson
mod5 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6  + agec5_24_years + agec25_44_years +
                                      agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older  + education_recode2 , 
            data=ds5, family='poisson')

#Negative binomial
# mod2 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6 + agec5_24_years + agec25_44_years +
#                                       agec45_64_years + agec65_74_years + agec75_84_years +  agec85_years_and_older, 
#             data=ds2)

ds5 <- cbind.data.frame(ds5,pred_interval(mod=mod5,model.type='poisson',ds.glm=ds5))

ggplot(data=ds5, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
  facet_wrap(education_recode~agec, scales='free', nrow=2)+
      geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') 

ggplot(data=ds5[ds5$year==2020,], aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
  facet_wrap(education_recode~agec, scales='free', nrow=2)+
      geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') 


ds5 <- ds5 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds5, aes(x=date, y=rr_median)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
  ylim(0.1,5) +
  geom_hline(yintercept=1, lty=2, col='red') +
    geom_vline(xintercept=as.Date('2020-03-01'), lty=2, col='gray80') +

  facet_wrap(education_recode~agec, nrow=2)
```

## Get latest data from CDC WONDER

```{r}
# dbnamelookup <- wonderapi::show_databases()
# 
# mylist <- list(list("Group Results By", "D176.V1-level1"), #year
#                list("And By", "D176.V1-level2"), #month
#                list("And By", "D176.V13-level3"),#by multiple cause of death
#                list("Month", "2") )  #subset MCD to those including A40-A41 
# 
# 
# #V_D176.V5 limit by multiple cause of death subchapter
# mydata2 <- getData("Provisional Multiple Cause of Death", mylist)

#Stratify by Year, month, ICD10; select MCD J13 and A40.3

mydata1 <- read.csv('./Data/Wonder/Provisional Mortality Statistics, 2018 through Last Month 2022_8_25.txt', header=T, sep="\t")

mydata2 <- mydata1 %>%
  mutate(date= as.Date(paste0(substr(Month,1,3),substr(Year,1,4), '01'), '%b%Y%d'), month=lubridate::month(date),
         year=as.factor(lubridate::year(date))) %>%
  filter(Multiple.Cause.of.death.Code !='') 


ggplot(mydata2, aes(x=month, y=Deaths, group=year, color=year)) +
  geom_line()+
  theme_classic()+
  ylim(0, NA) +
  facet_wrap(~Multiple.Cause.of.death.Code)
  

```

Market basket analysis
```{r}

key.vars <- pneumo.deaths %>%
    filter(year<2020) %>%
  dplyr::select(starts_with('icd') , starts_with('age_group'),starts_with('education_recode'),starts_with('sex')) %>%
  mutate(id=row_number())


pneumo.deaths.m <- reshape2::melt(key.vars, id.vars=c('id')) %>%
  filter(!is.na(value)) %>%
  mutate(variable=gsub('[0-9]+', '', variable), 
 item=paste(variable, value, sep='_')) %>%
  dplyr::select(id, item)

write.csv(pneumo.deaths.m,'./Raw Data/transacations.csv',row.names =F)

order_trans <- read.transactions(
        file = './Raw Data/transacations.csv',
        format = "single",
        sep = ",",
        cols=c(1,2),
        rm.duplicates = T
)


 itemFrequencyPlot(order_trans,topN=20,type="absolute")

 rules <- apriori(order_trans,parameter=list(supp=0.001,conf = 0.5),appearance = list(default="lhs",rhs="icd_J13"))

 summary(rules)

rules<-sort(rules, by="lift", decreasing=TRUE,)


options(digits=2)
inspect(rules[1:100])

plot(rules, method="graph", shading='lift')
```
2020 only
```{r}

key.vars2020 <- pneumo.deaths %>%
  filter(year==2020) %>%
  dplyr::select(starts_with('icd') , starts_with('age_group'),starts_with('education_recode'),starts_with('sex')) %>%
  mutate(id=row_number())


pneumo.deaths.m2020 <- reshape2::melt(key.vars2020, id.vars=c('id')) %>%
  filter(!is.na(value)) %>%
  mutate(variable=gsub('[0-9]+', '', variable), 
 item=paste(variable, value, sep='_')) %>%
  dplyr::select(id, item)

write.csv(pneumo.deaths.m2020,'./Raw Data/transacations2020.csv',row.names =F)

order_trans2020 <- read.transactions(
        file = './Raw Data/transacations2020.csv',
        format = "single",
        sep = ",",
        cols=c(1,2),
        rm.duplicates = T
)


 itemFrequencyPlot(order_trans2020,topN=20,type="absolute")

 rules2020 <- apriori(order_trans2020,parameter=list(supp=0.001,conf = 0.5),appearance = list(default="lhs",rhs="icd_J13"))

 summary(rules2020)

rules<-sort(rules2020, by="lift", decreasing=TRUE,)


options(digits=2)
inspect(rules[1:100])

plot(rules2020, method="graph", shading='lift')
```
```


```{r}

```





