---
title: "Analysis of pneumococcal deaths in the US"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F)

#install.packages("groundhog")
#library("groundhog")

library(dplyr)
library(stringr)
library(lubridate)
library(ggplot2)
library(MASS)
library(plotly)
library(arules)
library(usdata)
library(covidcast)
library(arulesViz)
library(directlabels)
library(arrow)
library(duckdb)
library(Rtsne)
library(cdlTools)
library(readr)
library(reshape2)
library(tidyr)
library(pbapply)
library(parquetize)
library(gridExtra)

library(DBI)

#groundhog.library(pkgs, "2022-04-21")

#groundhog.library('parquetize','2022-04-22')

source('./R/death_plot_fun.R')
source('./R/pred_interval.R')

source('./R/Format_US_Population.R')

#https://github.com/socdataR/wonderapi
#devtools::install_github("socdataR/wonderapi", build_vignettes = F, force=T)
#library(wonderapi)

```

Parquet file created with './R/Read_mortality_tapes_nongeo_parquet.R'

Following CDC wonder: 6 race categories - American Indian or Alaska Native (AIAN); Asian; Black or African American; More than one race; Native Hawaiian or Other Pacific Islander (NHOPI); White.

```{r}
pneumo.deaths.in <-  open_dataset("R:/NCHS_mortality_public/parquet3", format = "parquet")  %>%
  filter(grepl("J13|A39|J181|A403|B953|G001", all_icd)) %>%
 collect() %>%
  mutate( 
   a403 = 1*grepl('A403',all_icd),
    j13 = 1*grepl('J13', all_icd),
    j13_prim = substr(all_icd, 1,3) =='J13',
    b953 = 1*grepl('B953', all_icd),
    g001 = 1*grepl('G001', all_icd),
    pneumococcal =  1*((a403 + j13 + b953 + g001 )>0)  ,
   sepsis=grepl('A40', all_icd) |grepl('A419', all_icd), #streptococcal or unspecified sepsis codes
   meningitis=grepl('G00',all_icd),
   ipd = if_else(pneumococcal==1 & (sepsis==1|meningitis==1),1,0),
    covid = grepl('U071', all_icd),
    j154 =1 *grepl('J154', all_icd), #streptococcal pneumonia
    j181 =1 *grepl('J181', all_icd) , #unspecified lobar pneumonia
    j12_j18 = grepl('J12', all_icd) + grepl('J13', all_icd) + grepl('J14', all_icd) + grepl('J15', all_icd) + grepl('J16', all_icd) + grepl('J17', all_icd) + grepl('J18', all_icd),
   j12_j18 = if_else(j12_j18>1,1,j12_j18),
    streptococcal_sepsis = 1* ( grepl('A408',all_icd) +  
                                  grepl('A409',all_icd) )>0,
   pneumococcal_pneum = if_else( (j13==1 | (pneumococcal==1 & j12_j18==1)) ,1,0 ),
   pneumococcal_pneum_no_ipd = if_else(ipd==0,pneumococcal_pneum,0),
   agec = if_else(agey>=0 & agey<5,1,
                   if_else(agey>=5 & agey<25,2,
                   if_else(agey>=25 & agey<45,3,
                   if_else(agey>=45 & agey<65,4,
                   if_else(agey>=65 & agey<75,5,
                   if_else(agey>=75 & agey<85,6,
                   if_else(agey>=85 & agey<120,7,
                           999)))))))
                               ) %>%
 # filter(pneumococcal==1 | j181==1 | j154==1)
  filter(pneumococcal==1  & agec != 'missing') %>% #filter non-pneumo codes
  mutate(agec = factor(agec, c(1,2,3,4,5,6,7),c("Under 5 Years",'5-24 years', "25-44 years","45-64 years","65-74 years" ,"75-84 years","85 years and older")),
         race_recode_new = factor(race_recode_new, levels=c(1,2,3,4,5,999), c('White','Black','Hispanic','American Indian', 'Asian/Pacific Islanders','Other')) 
  )
  
table( pneumo.deaths.in$j12_j18)
table( pneumo.deaths.in$pneumococcal_pneum)
table( pneumo.deaths.in$ipd)
table( pneumo.deaths.in$pneumococcal_pneum_no_ipd)

table( pneumo.deaths.in$ipd,pneumo.deaths.in$pneumococcal_pneum_no_ipd)
table( pneumo.deaths.in$pneumococcal)

```


```{r}



deaths2019 <- pneumo.deaths.in %>% filter(year==2019) %>% summarize(N=n())

pneumo.deaths.in %>%
  group_by(agec,year) %>%
  filter(a403==1 & !is.na(agec)) %>%
  summarize(N=n())

pneumo.deaths.adult <- pneumo.deaths.in %>%
  mutate(one=1) %>%
  filter(agey>=25 & !is.na(agec)) %>%
  mutate(agec=droplevels(agec))

```
Popsize data
```{r}
#Format_US_Population_wonder()
pop1 <- readRDS('./Data/pop_interpol.rds')

pop2 <- pop1 %>%
  group_by(date, agec, race_recode) %>%
  summarize(pop.interpol= sum(pop.interpol)) %>%
  mutate(race_recode_new= factor(race_recode,levels=c(1,2,3,4,5), c('White','Black','Hispanic','American Indian', 'Asian/Pacific Islanders')),
         year=year(date),
         month=month(date))


# pop2.sex <- pop1 %>%
#   group_by(date, agec, sex) %>%
#   summarize(pop.interpol= sum(pop.interpol))
```




```{r}
p1_ipd <-  death_plot_fun(disease='ipd',reg.form='N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6',year,month )

p1_ipd$plot1 +
  ggtitle('IPD')


p1_pneumococcal_pneum_no_ipd <- death_plot_fun(disease='pneumococcal_pneum_no_ipd',reg.form='N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6',year,month )

p1_pneumococcal_pneum_no_ipd$plot1 +
  ggtitle('Non-invasive pneumococcal pneumonia')

```
Time series plots

```{r}
p1_ipd$ts.plot +
  ggtitle('IPD')

p1_pneumococcal_pneum_no_ipd$ts.plot +
  ggtitle('Non-invasive pneumococcal pneumonia')
```


### Stratified by age category

```{r, fig.width=12, fig.height=6}
p1_ipd <-  death_plot_fun(disease='ipd',reg.form='N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6 +agec',year,month,agec )

p1_ipd$plot1 +
  facet_wrap(~agec)+
  ggtitle('IPD')


p1_pneumococcal_pneum_no_ipd <- death_plot_fun(disease='pneumococcal_pneum_no_ipd',reg.form='N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6+agec',year,month,agec )

p1_pneumococcal_pneum_no_ipd$plot1 +
    facet_wrap(~agec)+
  ggtitle('Non-invasive pneumococcal pneumonia')
```


## Question 1: Did the rate of death due to pneumococcus change during the pandemic?

Fit model to 2014-2019 data, extrapolate to 2020 and beyond

IPD
```{r}
ds1 <- p1_ipd$ts %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-')),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         N_deaths_pre= if_else(date >= '2020-03-01', NA_real_,N_deaths),
         log_offset=log(1))

#Poisson
mod1 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6,
            data=ds1, family='poisson')

#Negative binomial
# mod1 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6,
#             data=ds1)

ds1 <- cbind.data.frame(ds1,pred_interval(mod=mod1,model.type='poisson'))

p1_a <- ggplot(data=ds1, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2)+
    geom_line(aes(x=date, y=pred_median), col='red', lty=2, alpha=0.2)+

    geom_vline(xintercept= as.Date(c('2020-03-01','2021-08-01')), lty=3, col='gray') +
    annotate("text", x=as.Date(c('2020-04-01','2021-09-01')), y=175, label=c('March 2020', 'August 2021'), angle=90, col='gray')+
  ylab('N deaths')+
  ggtitle('Invasive pneumococcal disease')
p1_a

ggplotly(p1_a)


ds1 <- ds1 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds1, aes(x=date, y=rr_median)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
 # ylim(0.5,2.2) +
  coord_cartesian(ylim = c(0.5, 2.2))+
  geom_hline(yintercept=1, lty=2, col='red')+    
  geom_vline(xintercept= as.Date(c('2020-03-01','2021-08-01')), lty=3, col='gray') +
    annotate("text", x=as.Date(c('2020-04-01','2021-09-01')), y=1.5, label=c('March 2020', 'August 2021'), angle=90, col='gray')

```

```{r}
p3_ipd <-  death_plot_fun(disease='ipd',reg.form='N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6 +race_recode_new',year,month,race_recode_new )


p3_ipd$plot1 +
  facet_wrap(~race_recode_new)+
  ggtitle('IPD')


p3_pneumococcal_pneum_no_ipd <- death_plot_fun(disease='pneumococcal_pneum_no_ipd',reg.form='N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6+race_recode_new',year,month,race_recode_new )

p3_pneumococcal_pneum_no_ipd$plot1 +
    facet_wrap(~race_recode_new)+
  ggtitle('Non-invasive pneumococcal pneumonia')
```

## Question 1: Did the rate of death due to pneumococcus change during the pandemic?




### How did the rate change by age group or ethnic group?



## How did rates change by education

Occupation--86% missing

```{r}
#unique(pneumo.deaths$occupation)

#sum(is.na(pneumo.deaths$occupation))/nrow(pneumo.deaths)
```

Education--2.5% missing

Education (1989 revision)
00 No formal education
00-08Years of elementary school
09 1 year of high school
10 2 years of high school
11 3 years of high school
12 4 years of high school
13 1 year of college
14 2 years of college
15 3 years of college
16 4 years of college
17 5 or more years of college
99 Not stated 

Education (2003 revision)
Field is blank for registration areas that are using the 1989 revision format of the item.
1 ... 8th grade or less
2 ... 9 - 12th grade, no diploma
3 ... high school graduate or GED completed
4 ... some college credit, but no degree
5 ... Associate degree
6 ... Bachelor’s degree
7 ... Master’s degree
8 ... Doctorate or professional degree
9 ... Unknown 

Education reporting flag
0 1989 revision of education item on certificate
1 2003 revision of education item on certificate
2 no education item on certificate

education_recode: 1=HS or less; 2= some college or more; 99=missing

```{r, fig.width=6, fig.height=4}

pneumo.deaths.adult <- pneumo.deaths.adult %>%
  mutate(education_hs = if_else(education1989 %in% c('00','01','02','03','04','05','06','07','08','09','10','11','12') |education2003 %in% c('1','2','3') ,1,0 ),
         education_college = ifelse( education1989 %in% c('13','14','15','16','17') | education2003 %in% c('4','5','6','7','8'),1,0),
         education_missing= ifelse( education1989 %in% c('99') | education2003 %in% c('9'),1,0),
         education_recode = ifelse(education_hs==1,1, ifelse(education_college==1,2,99) )
       
) %>%
  filter(education_recode %in% c(1,2)) %>%
  mutate(education_recode=as.factor(education_recode)
)
  



p4_ipd <-  death_plot_fun(disease='ipd',reg.form='N_deaths_pre ~ education_recode*(1+ index+ sin12+cos12 +sin6 +cos6)',year,month,education_recode )

p4_ipd$plot1 +
  facet_wrap(~education_recode)+
  ggtitle('IPD')


p4_pneumococcal_pneum_no_ipd <- death_plot_fun(disease='pneumococcal_pneum_no_ipd',reg.form='N_deaths_pre ~ education_recode*(1 + index+ sin12+cos12 +sin6 +cos6)',year,month,education_recode )

p4_pneumococcal_pneum_no_ipd$plot1 +
    facet_wrap(~education_recode)+
  ggtitle('Non-invasive pneumococcal pneumonia')
```

```{r, fig.width=6, fig.height=4}

p4_ipd$ts.plot +
    facet_wrap(~education_recode, scales='free') +
   ggtitle('IPD')


p4_pneumococcal_pneum_no_ipd$ts.plot +
    facet_wrap(~education_recode, scales='free') +
   ggtitle('Non-invasive pneumococcal pneumonia') #+
  #  annotate("text", x=as.Date(c('2020-04-01')), y=50, label='March 2020', angle=90, col='gray')+
  #   annotate("text", x=as.Date(c('2021-09-01')), y=50, label= 'August 2021', angle=90, col='gray')

```



## ABCs deaths from BactFacts (without serotype)
through 2021

```{r, fig.width=6, fig.height=3}
# d1 <-read.csv('https://data.cdc.gov/api/views/en3s-hzsr/rows.csv?accessType=DOWNLOAD')
# saveRDS(d1, './Data/abcs_data.rds')

death_abc <- readRDS('./Data/abcs_data.rds') %>%
  filter(Topic=='Death Rates' & ViewBy2 !='Overall' & ViewBy=='Age')%>%
  rename(agec=ViewBy2,death_rate=Value)

case_abc <- readRDS('./Data/abcs_data.rds') %>%
  filter(Topic=='Case Rates' & ViewBy2 !='Overall' & ViewBy=='Age') %>%
  rename(agec=ViewBy2,case_rate=Value)

cfr <- case_abc %>%
  left_join(death_abc, by=c('Year','agec')) %>%
  mutate(cfr= death_rate/case_rate)


cdc1 <- death_abc %>%
  filter(Year>=2014) %>%
ggplot( aes(x=Year, y=log(death_rate), color=agec)) +
         geom_line() +
  geom_point()+
         theme_classic() +
  xlim(2014, 2022.5)+
  geom_vline(xintercept=2019.5, lty=3) +
    geom_dl(aes(label = agec), method = "last.qp", cex=3) +
    # Remove legend & adjust margins to give more space for labels
    # Remember, the margins are t-r-b-l
    theme(legend.position = 'none',
          plot.margin = margin(0.1, 3, 0.1, 0.1, "cm")) +
    coord_cartesian(clip = 'off') 

cdc2 <- case_abc %>%
  filter(Year>=2014) %>%
ggplot( aes(x=Year, y=log(case_rate), color=agec)) +
         geom_line() +
  geom_point()+
         theme_classic() +
  xlim(2014, 2022.5)+
  geom_vline(xintercept=2019.5, lty=3) +
    geom_dl(aes(label = agec), method = "last.qp", cex=3) +
    # Remove legend & adjust margins to give more space for labels
    # Remember, the margins are t-r-b-l
    theme(legend.position = 'none',
          plot.margin = margin(0.1, 3, 0.1, 0.1, "cm")) +
    coord_cartesian(clip = 'off') 


cdc3 <-cfr %>%
  filter(Year>=2014) %>%
ggplot( aes(x=Year, y=(cfr), color=agec)) +
         geom_line() +
  geom_point()+
         theme_classic() +
  xlim(2014, 2022.5)+
  geom_vline(xintercept=2019.5, lty=3) +
    geom_dl(aes(label = agec), method = "last.qp", cex=3) +
    # Remove legend & adjust margins to give more space for labels
    # Remember, the margins are t-r-b-l
    theme(legend.position = 'none',
          plot.margin = margin(0.1, 3, 0.1, 0.1, "cm")) +
    coord_cartesian(clip = 'off') 

cdc1
cdc2
cdc3
```

## Get latest data from CDC WONDER
https://wonder.cdc.gov/controller/datarequest/D176
group by year, month, multiple cause of death MCD J13, A40.3
--IMPORTANT: it appears if you try to select multiple MCD codes and you stratify by MCD, it does weird things; o 1 at a time.
```{r}
#Note this first one is NOT reliable because it is stratifying by multiple MCD categories
mydata1 <- read.csv('./Data/Wonder/Provisional Mortality Statistics, 2018 through Last Week 2023_10_17.txt', header=T, sep="\t")

mydata2 <- mydata1 %>%
  mutate(date= as.Date(paste0(substr(Month,1,3),substr(Year,1,4), '01'), '%b%Y%d'), month=lubridate::month(date),
         year=as.factor(lubridate::year(date))) %>%
  filter(Multiple.Cause.of.death.Code !='') %>%
  rename(N_deaths=Deaths, cause=Multiple.Cause.of.death.Code)



##test1a: FINAL data; just MCD=J13
mydata_test1a <- read.csv('./Data/Wonder/Multiple Cause of Death, 2018-2021, Single Race.txt', header=T, sep="\t")

mydata_test2a <- mydata_test1a %>%
  mutate(date= as.Date(paste0(substr(Month,1,3),substr(Year,1,4), '01'), '%b%Y%d'), month=lubridate::month(date),
         year=as.factor(lubridate::year(date))) %>%
  filter(Multiple.Cause.of.death.Code !='') %>%
  rename(N_deaths_a=Deaths, cause=Multiple.Cause.of.death.Code) %>%
  dplyr::select(date, N_deaths_a)

##test1a: FINAL data; just MCD
mydata_test1a <- read.csv('./Data/Wonder/Multiple Cause of Death, 2018-2021, Single Race.txt', header=T, sep="\t")

##test1b: provisional data; just MCD=J13
mydata_test1b <- read.csv('./Data/Wonder/Provisional Mortality Statistics, 2018 through Last Week_j13_2023_10_18.txt', header=T, sep="\t")

mydata_test2b <- mydata_test1b %>%
  mutate(date= as.Date(paste0(substr(Month,1,3),substr(Year,1,4), '01'), '%b%Y%d'), month=lubridate::month(date),
         year=as.factor(lubridate::year(date))) %>%
  filter(Multiple.Cause.of.death.Code !='') %>%
  rename(N_deaths_b=Deaths, cause=Multiple.Cause.of.death.Code) %>%
    dplyr::select(date, N_deaths_b) %>%
  mutate(month=month(date),
         year=year(date),
         epiyr= as.factor(if_else(month>=7, year, year-1)),
         month=factor(as.character(month), levels=c('7','8','9','10','11','12','1','2','3','4','5','6'))
         ) 

# mydata_test2b %>%
#   dplyr::select(date, N_deaths_b) %>%
#  left_join(mydata_test2a, by='date') %>%
#   left_join(mydata_test2b, by='date')
  
p1 <- ggplot(mydata_test2b, aes(x=month, y=N_deaths_b, group=epiyr, color=epiyr)) +
  geom_line()+
  theme_classic()+
  ylim(0, NA) 

p1

ggplotly(p1)
```

Fit model to 2014-2019 data, extrapolate to 2020 and beyond

```{r}
##NCHS data:
p1 <-  death_plot_fun(disease='pneumococcal',year,month, j13 )

ds1_nchs <- pneumo.deaths %>%
  filter(j13==1) %>%
  group_by(year, month) %>%
    group_by( year, month ) %>%
    summarize(N_deaths=n()) %>%
    ungroup() %>%
    mutate(year=as.factor(year),month=as.numeric(as.character(month))) %>%
    tidyr::complete( year, month, fill=list(N_deaths=0) ) %>%
  mutate(date=as.Date(paste(year, month, '01', sep='-'))) %>%
  select(N_deaths, date) %>%
  rename(N_deaths_nchs=N_deaths) %>%
  group_by(date) %>%
  summarize(N_deaths_nchs=sum(N_deaths_nchs))


#WONDER DATA
ds1 <- mydata_test2b %>%
           rename(N_deaths_wonder=N_deaths_b) %>%
           full_join(ds1_nchs, by='date') %>%
    mutate(N_deaths= if_else( !is.na(N_deaths_wonder), N_deaths_wonder, N_deaths_nchs ),
         qtr=as.factor(lubridate::quarter(date)),
         index= interval('2014-01-01', date) %/% months(1) ,
         sin12 = sin(2*pi*index/12),
         cos12 = cos(2*pi*index/12),
         sin6 = sin(2*pi*index*2/12),
         cos6 = cos(2*pi*index*2/12),
         year=as.factor(year(date)),
         N_deaths_pre= if_else(date>='2020-03-01', NA_real_,N_deaths),
         log_offset=log(1)) 

ggplot(ds1, aes(x=date, y=N_deaths_wonder/N_deaths_nchs)) +
  geom_line()+
  #geom_line(aes(x=date, y=N_deaths_nchs), col='red')+
  theme_classic()+
  ggtitle('Comparison of final and provisional NCHS data')

#Poisson
mod1 <- glm(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6,
            data=ds1, family='poisson')

#Negative binomial
# mod1 <- glm.nb(N_deaths_pre ~ index+ sin12+cos12 +sin6 +cos6,
#             data=ds1)

ds1 <- cbind.data.frame(ds1,pred_interval(mod=mod1,model.type='poisson'))

p1 <- ggplot(data=ds1, aes(x=date, y=N_deaths)) +
  geom_line() +
  theme_classic()+
    geom_ribbon(aes(ymin=pred_lcl, ymax=pred_ucl), alpha=0.2) +
    geom_vline(xintercept= as.Date(c('2020-03-01','2021-08-01')), lty=3, col='gray') +
    annotate("text", x=as.Date(c('2020-04-01','2021-09-01')), y=175, label=c('March 2020', 'August 2021'), angle=90, col='gray') +
    #geom_line(aes(x=date, y=N_deaths_nchs),col='green') +
  #  geom_line(aes(x=date, y=N_deaths_wonder),col='blue') +
  ggtitle('Observed vs expected, adding CDC WONDER (CAUTION)')
p1
ggplotly(p1)

ds1 <- ds1 %>%
  mutate(rr_median=N_deaths/pred_median, rr_lcl=N_deaths/pred_lcl, rr_ucl=N_deaths/pred_ucl)


ggplot(data=ds1, aes(x=date, y=rr_median)) +
  geom_line() +
    theme_classic()+
    geom_ribbon(aes(ymin=rr_lcl, ymax=rr_ucl), alpha=0.2)+
 # ylim(0.5,2.2) +
  coord_cartesian(ylim = c(0.5, 2.2))+
  geom_hline(yintercept=1, lty=2, col='red')

```

age and race both have too many suppressed rows to make it interpretable from wonder
```{r, eval=F}

mydata1_race <- read.csv('./Data/Wonder/Provisional Mortality Statistics, 2018 through Last Month 2023_03_17 race.txt', header=T, sep="\t")

```

## ABCs case data, with serotype (no denominator)

```{r}
abc1 <- read.csv('https://data.cdc.gov/api/views/qvzb-qs6p/rows.csv?accessType=DOWNLOAD') %>%
  rename(agegrp=Age.Group..years.,
         st=IPD.Serotype,
         N_IPD=Frequency.Count) %>%
  group_by(agegrp, Year) %>%
  summarize(N_IPD= sum(N_IPD))

#saveRDS(abc1,'./Data/ABC_2016_2021.rds')


```

## Factors associated with pandmeic period
## Binary ICD variables

Create a vector with each of the unique ICD-10 codes (1827 total)
**or instead break into CCSR categories https://rdrr.io/github/HunterRatliff1/hcup/man/ccsr_dx.html **

```{r}

pneumococcal_deaths <- pneumo.deaths %>%
  filter(pneumococcal==1) %>%
  mutate(date = as.Date(paste(year,month,'01', sep='-')),
         education_hs = if_else(education1989 %in% c('00','01','02','03','04','05','06','07','08','09','10','11','12') |education2003 %in% c('1','2','3') ,1,0 ),
         education_college = ifelse( education1989 %in% c('13','14','15','16','17') | education2003 %in% c('4','5','6','7','8'),1,0),
         education_missing= ifelse( education1989 %in% c('99') | education2003 %in% c('9'),1,0),
         education_recode = as.factor(ifelse(education_hs==1,1, ifelse(education_college==1,2,99) ))
          ) %>%
  separate(all_icd, paste0('icd',1:21), sep='_', fill='right') %>%
      dplyr::select(starts_with('icd') , starts_with('agec'),starts_with('education_recode'),starts_with('sex'), race_recode_new,date) %>%
    mutate(id=row_number()) %>%
    mutate(across(starts_with("icd"), ~substr(.,1,3))) %>%
   mutate(all_icd = paste(icd1,icd2,
                                 icd3,icd4,
                                 icd5,icd6,
                                 icd7,icd8,
                                 icd9,icd10,
                                 icd11,icd12,
                                 icd13,icd14,
                                 icd15,icd16,
                                 icd17,icd18,
                                 icd19,icd20,
                                 icd21, sep='_')
   ) %>%
  filter(!is.na(agec) & !is.na(sex) & !is.na(education_recode) & !is.na(race_recode_new))

all_icd.3digit2 <- unique(unlist(str_split(paste(pneumococcal_deaths$all_icd, collapse='_'),'_')))

big.mat_full <- pbsapply(all_icd.3digit2, function(X) var1 = grepl(X, pneumococcal_deaths$all_icd) , simplify='array')

big.mat <- big.mat_full %>% 
  as.data.frame() %>% 
  distinct() %>%
  as.matrix() 

big.mat.t <- t(big.mat) %>% 
  as.data.frame() %>% 
  distinct() %>%
  as.matrix() 

#tsne_icd <- Rtsne(big.mat)

#tsne_icd2 <- Rtsne(big.mat.t) #plots the ICD codes
#try again, but remove outcome codes from matrix
big.mat2a <- big.mat
row.names(big.mat2a) <- 1:nrow(big.mat)
   
#B953 not found
big.mat2b <- big.mat2a %>%
  as.data.frame() %>%
  dplyr::select(-J13,-A403, -J181, -G001) %>%
  distinct() %>%
  as.matrix()

ids.kept <- row.names(big.mat2b)

tsne_icd3 <- Rtsne(big.mat2b)

label.mat <- big.mat2a %>%
  as.data.frame() %>%
  filter(row.names(.) %in% ids.kept) 

# Conversion of matrix to dataframe
tsne_plot <- data.frame(x = tsne_icd3$Y[,1],
                        y = tsne_icd3$Y[,2])
 
# Plotting the plot using ggplot() function
p1 <- ggplot2::ggplot(tsne_plot)+ 
  geom_point(aes(x=x,y=y, text=))
plotly::ggplotly(p1)

ggplot2::ggplot(tsne_plot)+ 
  geom_point(aes(x=x,y=y, color=label.mat[,'J13']))

ggplot2::ggplot(tsne_plot)+ 
  geom_point(aes(x=x,y=y, color=label.mat[,'F102'])) 


sort(-apply(big.mat,2,mean))[1:20]

#now can do logistic regression--what icd codes predict whether the person has A40.3 vs J13 vs J18.1
```

Ridge Regression: are there codes associated with pandemic period, or other characteristics?
```{r}
library(glmnet)

mod.ds <-  big.mat_full %>%
  as.data.frame() %>%
  mutate(Y = if_else(pneumococcal_deaths$date >= '2020-03-01',1,0))

x.covar <-   model.matrix( ~agec + sex +education_recode + race_recode_new  ,data=pneumococcal_deaths) 
x.covar <- x.covar[,-1]


X <- mod.ds %>% 
  dplyr::select(-Y) %>%
  as.matrix() %>%
  cbind(.,x.covar)

Y= mod.ds %>% 
  dplyr::select(Y) %>%
  pull(Y)

set.seed(123) 
cv.lasso <- cv.glmnet(x=X, y=Y, alpha = 1, family = "binomial")

# Fit the final model on the training data
mod1 <- glmnet(x=X, y=Y, alpha = 1, family = "binomial", lambda = cv.lasso$lambda.min)

coefs <- coef(mod1)[,1]

sort(coefs, decreasing=T)[1:100]
sort(coefs, decreasing=F)[1:100]

# download.file("https://github.com/cran/icd.data/raw/master/data/icd10_sub_chapters.RData", "icd10")
# load("icd10")
# saveRDS(icd10_sub_chapters,'./Data/icd10_subchaper.rds')
icd10_sub_chapters <- readRDS('./Data/icd10_subchaper.rds')
icd10_sub_chapters[["Human Immunodeficiency Virus [HIV] Disease"]]['end'] <- 'B24'
icd10_sub_chapters[["Diabetes Mellitus"]]['end'] <- 'E14'
icd10_sub_chapters[["Event Of Undetermined Intent"]]['end'] <- 'Y34'


covid <- list('covid'=c('start'='U07', 'end'='U07'))

icd10_sub_chapters <- c(icd10_sub_chapters, covid)
icd_mapping <- lapply(names(coefs), function(X){
       
    code.check <- sapply(1:length(icd10_sub_chapters), function(Y)
     {
           code_present <- X >=icd10_sub_chapters[[Y]][1] &  X <=icd10_sub_chapters[[Y]][2]
           code_name <- names(icd10_sub_chapters)[Y]
           c('icd'=X,'code_present'=code_present,'code_name'=code_name)
     })
   code.check <- as.data.frame(t(code.check))
  }
) %>%
  bind_rows() %>%
  filter(code_present.start==TRUE)

coefs.df <- cbind.data.frame('var'=names(coefs),'coef'=coefs) %>%
  left_join(icd_mapping, by=c('var'='icd')) %>%
  filter(var !='(Intercept)' & var !='NA') %>%
  arrange(coef) %>%
  mutate( code_name = if_else(is.na(code_name),var,code_name),
          coef=round(coef,2)) %>%
  dplyr::select(coef ,var,code_name ) %>%
  filter(abs(coef)>0.1)

write.csv(coefs.df,'./Data/ridge_regression.csv')
View(coefs.df)


```

CCSR instead!
```{r}
#install.packages("remotes")
#remotes::install_github("HunterRatliff1/hcup")
library(hcup)

pneumococcal_deaths <- pneumo.deaths %>%
  filter(pneumococcal==1) %>%
  mutate(date = as.Date(paste(year,month,'01', sep='-')),
         education_hs = if_else(education1989 %in% c('00','01','02','03','04','05','06','07','08','09','10','11','12') |education2003 %in% c('1','2','3') ,1,0 ),
         education_college = ifelse( education1989 %in% c('13','14','15','16','17') | education2003 %in% c('4','5','6','7','8'),1,0),
         education_missing= ifelse( education1989 %in% c('99') | education2003 %in% c('9'),1,0),
         education_recode = as.factor(ifelse(education_hs==1,1, ifelse(education_college==1,2,99) )),
         ID=row_number()
        ) %>%
  separate(all_icd, paste0('icd',1:21), sep='_', fill='right') %>%
  dplyr::select(ID, date,year, month, sex,education_recode, agec, race_recode_new, starts_with('icd') ) %>%
  reshape2::melt(., id.vars=c('ID','date','year', 'month', 'sex','education_recode', 'agec', 'race_recode_new')) %>%
filter(!is.na(value))


#test <- ccsr_dx(pneumococcal_deaths$value)
test <- classify_ccsr_dx1(pneumococcal_deaths$value,setting = "inpatient")

test2 <- lapply(test, function(X) paste(X, collapse='_')) %>% unlist()

pneumococcal_deaths2 <- pneumococcal_deaths %>%
  mutate(ccsr = test2 ) %>%
  separate(ccsr, paste0('ccsrcat',1:3), sep='_', fill='right') %>%
    dplyr::select(ID, year,date, month, sex,education_recode, agec, race_recode_new, starts_with('ccsrcat') ) %>%
  reshape2::melt(., id.vars=c('ID','year','date', 'month', 'sex','education_recode', 'agec', 'race_recode_new')) %>%
  filter(!is.na(value) & value !='NA') %>%
  dplyr::select(-variable) %>%
  rename(ccsr=value) %>%
  distinct() %>%
  group_by(ID, year,date, month, sex,education_recode, agec, race_recode_new) %>%
  summarize( ccsr = paste(ccsr, collapse='_')) %>%
  filter(!is.na(sex) & !is.na(education_recode) & !is.na(agec) & !is.na(race_recode_new))

all_ccsr <- unique(unlist(str_split(paste(pneumococcal_deaths2$ccsr, collapse='_'),'_')))

big.mat_full <- pbsapply(all_ccsr, function(X) var1 = grepl(X, pneumococcal_deaths2$ccsr) , simplify='array')




library(glmnet)

mod.ds <-  big.mat_full %>%
  as.data.frame() %>%
  mutate(Y = if_else(pneumococcal_deaths2$date >= '2020-03-01',1,0))

x.covar <-   model.matrix( ~agec + sex +education_recode + race_recode_new  ,data=pneumococcal_deaths2) 
x.covar <- x.covar[,-1]


X <- mod.ds %>% 
  dplyr::select(-Y) %>%
  as.matrix() %>%
  cbind(.,x.covar)

crcs_freq <- apply(X,2, sum)
X_filt <- X[,crcs_freq>=100]
crcs_freq_filt <- apply(X_filt,2, sum)

crcs_freq_filt_df <- cbind.data.frame('crcs'=names(crcs_freq_filt), crcs_freq_filt)


Y= mod.ds %>% 
  dplyr::select(Y) %>%
  pull(Y)

set.seed(123) 
cv.lasso <- cv.glmnet(x=X_filt, y=Y, alpha = 0, family = "binomial")

# Fit the final model on the training data
mod1 <- glmnet(x=X_filt, y=Y, alpha = 0, family = "binomial", lambda = cv.lasso$lambda.min)

coefs <- coef(mod1)[,1]

coef.df <- cbind.data.frame('coefs'=coefs, 'crcs_exp'=hcup::explain_ccsr(names(coefs)), 'crcs'=names(coefs) ) %>%
  left_join(crcs_freq_filt_df, by='crcs') %>%
  arrange(-coefs)


mod2 <- glmnet(x=X_filt, y=Y, alpha = 0, family = "binomial")



#descriptive hcup.data::CCSR_labels
table(mod.ds$END005)
table(mod.ds$END002)

table(mod.ds$END005, mod.ds$END002)
```

Market basket analysis
```{r}

key.vars <- pneumo.deaths %>%
    filter(year<2020) %>%
    mutate(education_hs = if_else(education1989 %in% c('00','01','02','03','04','05','06','07','08','09','10','11','12') |education2003 %in% c('1','2','3') ,1,0 ),
         education_college = ifelse( education1989 %in% c('13','14','15','16','17') | education2003 %in% c('4','5','6','7','8'),1,0),
         education_missing= ifelse( education1989 %in% c('99') | education2003 %in% c('9'),1,0),
         education_recode = ifelse(education_hs==1,1, ifelse(education_college==1,2,99) )
         ) %>%
   separate(all_icd, paste0('icd',1:21), sep='_', fill='right') %>%
    dplyr::select(starts_with('icd') , starts_with('agec'),starts_with('education_recode'),starts_with('sex')) %>%
  mutate(id=row_number())


pneumo.deaths.m <- reshape2::melt(key.vars, id.vars=c('id')) %>%
  filter(!is.na(value)) %>%
  mutate(variable=gsub('[0-9]+', '', variable), 
 item=paste(variable, value, sep='_')) %>%
  dplyr::select(id, item)

write.csv(pneumo.deaths.m,'./Data/transacations.csv',row.names =F)
```

```{r}
order_trans <- read.transactions(
        file = './Data/transacations.csv',
        format = "single",
        sep = ",",
        cols=c(1,2),
        rm.duplicates = T
)


 itemFrequencyPlot(order_trans,topN=20,type="absolute")

 rules <- apriori(order_trans,parameter=list(supp=0.01,conf = 0.5),appearance = list(default="lhs",rhs="icd_J13"))

 summary(rules)

rules1<-sort(rules, by="lift", decreasing=TRUE)


options(digits=2)
inspect(rules1[1:100])
inspectDT(rules1[1:1000])


rules2<-sort(rules, by="support", decreasing=TRUE)
inspect(rules2[1:100])


plot(rules1, method="graph", shading='lift', limit=100)
```


2020 only
```{r}

key.vars2020 <-  pneumo.deaths %>%
    filter(year==2020) %>%
    mutate(education_hs = if_else(education1989 %in% c('00','01','02','03','04','05','06','07','08','09','10','11','12') |education2003 %in% c('1','2','3') ,1,0 ),
         education_college = ifelse( education1989 %in% c('13','14','15','16','17') | education2003 %in% c('4','5','6','7','8'),1,0),
         education_missing= ifelse( education1989 %in% c('99') | education2003 %in% c('9'),1,0),
         education_recode = ifelse(education_hs==1,1, ifelse(education_college==1,2,99) )
         ) %>%
   separate(all_icd, paste0('icd',1:21), sep='_', fill='right') %>%
    dplyr::select(starts_with('icd') , starts_with('agec'),starts_with('education_recode'),starts_with('sex')) %>%
  mutate(id=row_number()) %>%
  mutate(across(starts_with("icd"), ~substr(.,1,3)))


pneumo.deaths.m2020 <- reshape2::melt(key.vars2020, id.vars=c('id')) %>%
  filter(!is.na(value)) %>%
  mutate(variable=gsub('[0-9]+', '', variable), 
 item=paste(variable, value, sep='_')) %>%
  dplyr::select(id, item)

write.csv(pneumo.deaths.m2020,'./Data/transacations2020.csv',row.names =F)

order_trans2020 <- read.transactions(
        file = './Data/transacations2020.csv',
        format = "single",
        sep = ",",
        cols=c(1,2),
        rm.duplicates = T
)


 itemFrequencyPlot(order_trans2020,topN=20,type="absolute")

 rules2020 <- apriori(order_trans2020,parameter=list(supp=0.01,conf = 0.5),appearance = list(default="lhs",rhs="icd_J13"))

 summary(rules2020)

rules<-sort(rules2020, by="lift", decreasing=TRUE)


options(digits=2)
inspect(rules[1:100])

plot(rules2020, method="graph", shading='lift', limit=100)
```


```{r, eval=F}
age_deaths <-  open_dataset("Y:/NCHS_mortality_GEO", format = "parquet")  %>%
mutate(age_detail_number = as.numeric(age_detail_number),

         agey= if_else(age_detail_class==1, age_detail_number,
              if_else(    age_detail_class==2,age_detail_number/12,
         if_else(age_detail_class==4,age_detail_number/365,
         if_else(age_detail_class==5,age_detail_number/365/24,
         if_else(age_detail_class==6,age_detail_number/365/24/60,99999))))), 
  agec = if_else(agey>=0 & agey<5,1,
                   if_else(agey>=5 & agey<25,2,
                   if_else(agey>=25 & agey<45,3,
                   if_else(agey>=45 & agey<65,4,
                   if_else(agey>=65 & agey<75,5,
                   if_else(agey>=75 & agey<85,6,
                   if_else(agey>=85 & agey<120,7,
                           999))))) ))
) %>%
  group_by(agec, year, month) %>%
  summarize(N_deaths=n()) %>% 
collect()
                                
```








